Probability of causal contact between interstellar civilizations through MonteCarlo simulations

### 2020-05-18 submit_v03/manuscript.tex
fin primera iteracion

In what follows, we outline the experimental setup adopted to estimate the probabilities of causal contacts and several derived quantities.
%
This is done in terms of three independent parameters, namely, the mean time span between the appearance of consecutive \cetis{}, $\left<\tau_a\right>$, their mean lifetime $\left<\tau_s\right>$, and the maximum distance a signal can be detected by another \ceti{} ($D_{max}$).
%
Intuitively, the probability of the existence of causal contacts between pairs of \cetis{} would be larger for smaller $\tau_a$ parameters, higher $\tau_s$ or higher $D_{max}$ parameter values.
%
We also propose theoretical distribution functions for both the lifespans ($\tau_s$) and the number of \cetis{} per unit time \citep{maccone_evolution_2014, Sotos_biotechnology_2019}.
%
The later is related to the time span between the appearance of consecutive \cetis{} (since when $\left<\tau_a\right>$ is shorter, it produces a greater density of \cetis{}).
%
The analytical expresiones for these distributions are set to a fixed law, as discussed in Sec.~\ref{SS_PDF_shape}.

@@

\begin{figure}[!t]
   \centering
   \includegraphics[width=0.5\textwidth]{F_scheme.pdf}
   \caption{
   %
Space--time diagrams showing a schematic representation of the
different stages in the development of causal contact node (\ceti{}).
%
Panel (a) represents the region $R$ in space and time which is causally
connected to the emitter (left vertex), following an "A" type event
("Awakening") in which the node acquires the communication capability.
%
The sphere of first contact (SFC) is a sphere centered on the emitter
that grows until its radius reaches the D$_{max}$ distance, in which
the power of a signal would equal the detectability threshold.
%
In the Figure this sphere is represented by the left triangle of the
region $R$.
%
The surface of last contact (SLC) is another sphere that grows from a
"D" event ("Doomsday").
%
The region which is causally connected to the emitter is then limited
by these two spheres, and has the shape of a sphere or of a spherical
shell, depending on the time.
%
The temporal intervals for
the communication between two \cetis{} are represented in he
panel (b).
%
The receiver \ceti{} E$_1$ can listen signals from the emiter \ceti{} E$_2$,
from the "Contact" event (t=C$_{12}$) up to the "Blackout" event (t=B$_{12}$).
%
Similarly, 
the receiver \ceti{} E$_2$ can listen signals from the emiter \ceti{} E$_1$,
from t=C$_{21}$ up to t=B$_{21}$. 
%
   } \label{F_scheme}
\end{figure}

@@

\ffn{1} Space—time diagrams, where time and space are represented on
the horizontal and vertical axes, respectively, are suitable tools to
describe the causal connections among different nodes.
%
We illustrate in the Fig.~\ref{F_scheme} the schematic representation
of the region causally connected to a given \ceti{}.
%
For the sake of simplicity, we show in the plot the light travel
distance, i.e., the distance traversed by any signal spreading from
the emitter at light speed (for example, any electromagnetic signal).
%
In this scheme, a light pulse would follow a trajectory represented by
a line at 45 degrees from the axes, given the units of the time and
space axes are years and light years, respectively.
%
Panel (a) of the Figure represents the region $R$ (shaded polygon)
which is causally connected to the emitter (left vertex).
%
This region develops after an event (hereafter dubbed A-type event) in
which the node acquires the communication capability, or becomes
''active''.
%
The sphere of first contact (SFC) is centred on the emitter,
represented in the Figure by the left angle of the region $R$.
%
This sphere grows until its radius reaches the D$_{max}$ distance, in
which it would be no longer detectable due to the decrease in the
energy per unit area, which falls under the assumed detectability
threshold.
%
Similarly, the surface of last contact (SLC) is another sphere that
grows from an event in which the nodes ceases to possess the
communication capability (D type event), and carries the last signal
produced by the emitter.
%
The region which is causally connected to the emitter is then limited
by these two spheres and therefore has the shape of a filled sphere
(before the D event) or of a spherical shell (after the D event).
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b) of the Figure.
%
In this scheme, the entire lifetime of a \ceti{} can be represented by
a polygon, as shown in the Figure.
%
The polygon is limited in the spatial axis by the maximum distance of
the signal $D_{max}$ and in the time axis by the time span between the
A-type event (concave vertex) and the D-type event (convex vertex).
%
This representation allows to visualize the important events that
result from the presence and communications of two \cetis{}.
%
The points in time where the \cetis{} acquire their communicating
capacity are dubbed ‘’Awakening’’ events ($A_1$ and $A_2$ in the
Figure).
%
Similarly, the instants in time where the \cetis{} lose their
communicating capacity are dubbed ‘’Doomsday’’ events ($D_1$ and $D_2$
in the Figure).
%
The points in space--time where the first contact is produced for each
one of the \cetis{}, are defined as ''Contact'' events, shown as $C_1$
and $C_2$.
%
Finally, the two points where the contact is lost for each one the
\cetis{}, are denominated ''Blackout'' events, presented as $B_1$ and
$B_2$ in the Figure.
%
The receiver node E$_1$ can listen signals from the emitter node
E$_2$, from the first contact event (t=C$_{12}$) until the last
one (t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$.
%
The time intervals for the open communication channel are determined
by the ''C'' and ''B'' type events.
%
It is noteworthy the fact that there is a time delay between the
contact events of the two \cetis{} involved in this analysis, and also
a time delay between the two blackout events.
%
Therefore, the time range when a bidirectional contact is possible
occurs between the maximum time of the contact points and the minimum
time between the blackout points.

@@

The leading front, or surface of first contact (SFC) grows from the central \ceti{} until it reaches the maximum distance $D_{max}$.

@@

This region will grow if the surface of first contact has not yet reached the maximum distance $D_{max}$, and will shrink after a D-type event until the surface of last contact reaches $D_{max}$, producing as a result the loss of all signal from the central \ceti{}.

@@

 \setlength{\tabcolsep}{10pt}
\begin{table*}
\centering
\begin{tabular}{cllll}
\hline
   \multicolumn{2}{l}{independent variables (hypothesis space)}
   &min. value&max. value&Nbins\\
\hline
   $\tau_{a}$ & Mean temporal separation between consecutive
   awakenings & 4000 yr & 200\,000 yr & 50\\ 
   $\tau_{s}$ & Mean lifetime of a \ceti{}
   & 10000 yr & 500\,000 yr& 50\\ 
   D$_{max}$ & Maximum reach of a message  &  \multicolumn{2}{l}{(500,
   1000, 10000, 20000, 40000, 80000) lyr} & 6  \\
\hline
   \multicolumn{2}{l}{fixed variables} & \multicolumn{2}{c}{assumptions} &value \\
\hline
   & statistical properties of all \cetis{} &equally distributed&&\\
   & Point process for the distribution in time & homogeneous &&\\
   f$_s$ & The scan of the sky & fully efficient&&1\\
   f$_p$ & panspermia or colonization &absent&&0\\
   & shape of the Galactic Habitable Zone & 2--dimensional ring &&\\
   R$_{GHZ}^{min}$   & Inner radius of the GHZ     & & &20\,000 lyr\\
   R$_{GHZ}^{max}$   & Outer radius of the GHZ     &&&60\,000 lyr\\
   t$_{max}$ & Time span of the simulation  &&&1.e6 yr\\
    & \multicolumn{3}{l}{number of random realizations for each point in the hypothesis space} &50\\
\hline
   \multicolumn{2}{l}{discrete events} &
   \multicolumn{3}{l}{affected variables}\\
\hline
   A event & Awakening: a \ceti{} starts its communication
   capabilities &Number of active \cetis{}\\
   B event & Blackout: the end of the communication channel stops
   &Number causal channels\\
   C event & Contact: a new causal contact is produced &Number of causal channels\\
   D event & Doomsday: a \ceti{} ends its communication capabilities&Number of active \cetis{}\\
\hline

@@

The temporal structure of the process is defined by two distribution parameters. One of them represent the mean time
interval that a \ceti{} can emit and receive signals, and the other the mean time interval between the emergence of
consecutive \cetis{}. The spatial structure of the simulation is given by the size and shape of the Galactic Habitable
Zone and the maximum distance a signal can travel to be detected (D$_{max}$). The parameters for the temporal
distributions also determine the spatial properties, since the density of active \cetis{} in the galaxy depend on these
two parameters. For example, a small $\tau_a$ and a large $\tau_s$ will produce a galaxy densely populated with
\cetis{}. Also, some hypotheses regarding the shapes of the distributions of the temporal parameters must be made in
order to complete the simulation. \citet{forgan_spatiotemporal_2011} argues that the times at which different
civilisations become intelligent follow a Gaussian distribution, and then the distribution of inter-arrival times is an
inverse exponential. We assume that the distribution of the times of A events is a stationary Poisson process, and then
the distribution of the times between the appearance of new consecutive \cetis{} is exponential. Regarding the duration
of a \ceti{}, we propose that the distribution of the duration of \cetis{} is a stationary exponential distribution.
There are no clear arguments to conclude a statistical law for the later distribution, so that we propose it as a
working hipothesis.  A theoretical statistical distribution of the lifespan of civilisations would rely preferentially
on the basis of the underlying astrophysical and biological processes \citep{balbi_impact_2018}. For example,
\citet{maccone_lognormals_2014} argues that this distribution should be a log normal.

@@

\begin{figure} \centering
   \includegraphics[width=0.5\textwidth]{F_number_of_contacts.pdf}
   \caption{
   %
Empirical cumulative distributions of the number contacts
for \cetis in four different samples (M5, M6, M7, M8) with two
values for the range of the signal reach: D$_{max}$=40000 ly (blue),  and 
D$_{max}$=80000 ly (orange).
%
Except for the models corresponding to a densely populated galaxy with
long--living \cetis, the \cetis which receive more than 10 contacts
are very unlikely.
%
See Table \ref{T_selected_models} for model descriptions.
%
   } \label{F_number_of_contacts}
\end{figure}

@@

In this Section we discuss the degree of complexity in the model, considering a compromise between the accuracy of the
model and the number of parameters that are free or with a high uncertainty.
%
Firstly, we emphasise that the odds of a causal contact between two \cetis{} should not be considered as the odds of a
contact between two intelligent civilisations, and in fact the latter could be much lesser than the former.
%
Indeed, in order to establish a contact between any two entities, a minimum degree of compatibility must be accomplished
without any previous agreement, making the possibility of a contact with a message that could be deciphered highly rare
\citep[see e.g. ][]{forgan_collimated_2014}.
%
Besides the trade--off between the simplicity and the complexity of the experiments, further analysis could be performed
following this framework in order to explore possible implications of the results for more detailed configurations.
%
For example, the communication method (isotropic, collimated, serendipitous) can affect the observables, making it
necessary to implement a correction factor.
%
Taking this into account, our results regarding the probability of causal contact should be considered as upper limits
for effective contacts, since they depend on the efficiency of both the emitter and the receiver to broadcast and scan
the sky for intelligent signals, respectively \citep{grimaldi_signal_2017}.
%
A correction by coverage ratio in the detection and by a targeting ratio in the emission could be easily implemented in
the simulation, although the effect of reducing the probability of contact is basically the product of the efficiency
ratios and thus such implementation is not necessary.
%
Therefore, the values of the probabilities could be modified by a constant correction factor equal to the combined
emission/reception efficiency \citep{smith_broadcasting_2009, anchordoqui_upper_2019, forgan_collimated_2014} or for
beam-like transmissions \citep{grimaldi_signal_2017}. 
%
Other considerations include the effects of alignments, the use of stars as sources or amplifiers \citep{Edmondson2003,
borra_searching_2012}, the nature of the message carrier.
%
Another improvement could be the use of a spatial distribution that resembles the spiral shape and the width of the disc
of the Galaxy.
%
Regarding the distributions of the parameters, different distribution functions for the mean lifetime of \cetis{} can be
implemented as alternatives.
%
Finally, regarding the degree of idealisation, the model admits different efficiencies for the sphere of the causal
region of each \ceti{}, featuring different searching strategies \citep{hippke_interstellar_2017}. 
%
It is also possible to consider that $D_{max}$ is different for different \cetis{}.
%
For example, a power law where powerful emissions are rare and low emissions are common could be an improvement to the
model.
%
In this work we chose not to implement this for the sake of simplicity.
%
Finally, the role of the message contents could influence on the lifespan of a \ceti{} that receives a message, although
the implementation of such behaviour would increase the number of free parameters and would be more speculative.

@@

A discrete event simulation is performed for a given model $M(\tau_a, \tau_s, D_{max})$ by keeping track of a set of variables that change each time an event happens. 
%
In addition, the model comprises elements that are fixed for all simulations, for example, the functional forms of the statistical distributions and the adopted values of particular variables (see Table XXX).
%
The main variables that follow the evolution of the simulation are: the positions of stars, which are sampled randomly within the GHZ; the time of the awakening of each \ceti{} (A event); and the time of the vanishing of each \ceti{} (D event). 
%
The variables that can be deduced from the previous ones include: the number of \cetis{} in casual contact with at least another \ceti{} at a given time; the number of \cetis{} as a function of time; the number of \cetis{} that receive a message at least one time; the number of \cetis{} that receive a message at least one time and successfully deliver an answer; and the number distribution of waiting time to receive a message.
%
All these quantities are updated each time one of the four events (A, B, C, D) occurs.

@@

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{F_never_contact.pdf}
   \caption{
   %
Rate of MPLs that never make contact (listening) as a
function of $\tau_a$ and $\tau_s$, for 
$D_{max}$=10000 (left panel),
$D_{max}$=40000 (middle panel), and
$D_{max}$=80000 (right panel)
%
   }
   \label{F_never_contact}
\end{figure*}

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{F_C_at_A.pdf}
   \caption{
   %
Rate of MPLs that listen at the moment of the
awakening, as a funcion of $\tau_a$ and $\tau_s$, for
$D_{max}$=10000 (left panel),
$D_{max}$=40000 (middle panel), and
$D_{max}$=80000 (right panel)
%
   }
   \label{F_C_at_A}
\end{figure*}

@@

% hearsay 
%
With the aim of making a reproducible project, we developed the tools
that allow to run and obtain the results shown in this work.
%
The simulations were implemented on a Python-3 code, dubbed \hs{},
which is publicly accessible through the GitHub
platform\footnote{https://github.com/mlares/simu\_contact} under the
MIT-license.
%
The project is in the process of registration with the ‘’Astrophysics
Source Code Library’’ \citep[ASCL, ][]{2015JORS....3E..15A,
2020ASPC..522..731A}
%
In order to achieve high performance we made use of vectorised methods
and array broadcasting properties provided by the NumPy package
\citep{oliphant_numpy_2006, vanderwalt_numpy_2011}
%
The combination of C-compiled Numpy routines for numerical
computations with a high-level language allows to obtain an easily
comprehensible and maintainable code.
%
From the viewpoint of a user, \hs{} is an object-oriented package that
exposes the main functionalities as classes and methods.
%
In the configuration step, the user prepares the set of simulation
parameters through an initialisation file.
%
Since configuration files and simulation results are persistent, it is
straightforward to keep track of different experiments and the
experiments can be revisited easily.
%
An in-depth description of the methods can be found in the
documentation, which is automatically generated from \hs{} docstrings
and made public in the read-the-docs service.

@@

\setlength{\tabcolsep}{10pt}
\begin{table*}
\centering
\begin{tabular}{ccccc}
\hline
   Model subset & D$_{max}$ [lyr] & $\tau_a$ interval [yr] & $\tau_s$ interval 
   [yr]& description  \\
\hline
M1 & 40000 & [10000, 30000]   & [10000, 50000]   &dense, short lifetime\\
M2 & 40000 & [170000, 190000] & [10000, 50000]   &sparse, short lifetime\\
M3 & 40000 & [10000, 30000]   & [400000, 440000] &dense, long lifetime \\
M4 & 40000 & [170000, 190000] & [400000, 440000] &sparse, long lifetime\\
%
M5 & 80000 & [10000, 30000]   & [10000, 50000]   &dense, short lifetime\\
M6 & 80000 & [170000, 190000] & [10000, 50000]   &sparse, short lifetime\\
M7 & 80000 & [10000, 30000]   & [400000, 440000] &dense, long lifetime \\
M8 & 80000 & [170000, 190000] & [400000, 440000] &sparse, long lifetime\\
%
\hline
\end{tabular}
\caption{Selected models to analyze the behavior of simulation outputs
   and their dependence on simulation parameters.}
\label{T_selected_models}
\end{table*}

@@

We implemented the simulation of a regular grid of models varying over
the hypothesis space, which covers 15000 different models.
%
For each model, we simulated 50 realizations with different random
seeds, in order to perform a MonteCarlo estimation of the variance.
%
The parameters for the temporal aspects of the simulation (the mean
waiting time for the next awakening, $\left<\tau_a\right>$, and the
mean lifetime, $\left<\tau_s\right>$) cover the ranges 4000-200000 yr
and 10000-500000 yr, respectively, with a regular partition of 50
values for each parameter.
%
For the D$_{max}$ parameter, we chose the values 500, 1000, 10000,
20000, 40000 and 80000 lyr.
%
This setup makes a total of 750000 simulations, each covering a time
range of one million years.
%
\ttn{1}
%
In Table \ref{T_simu_hypotheses} we show the three variable
parameters, the ranges of their values and the number of bins that
have been explored in the numerical experiments.
%
We also show the set of parameters that take part in the simulation,
their values and the hypotheses that define the runs of the
simulations.

>>>>

In what follows, we outline the experimental setup adopted to estimate the probabilities of causal contacts and several derived quantities.
%
This is done in terms of three independent parameters, namely, the
mean time span between the appearance of consecutive \cetis{},
$\left<\tau_a\right>$, their mean lifetime $\left<\tau_s\right>$, and
the maximum distance a signal can be detected by another \ceti{}
($D_{\rm max}$).
%
Intuitively, the probability of the existence of causal contacts
between pairs of \cetis{} would be larger for smaller $\tau_a$
parameters, higher $\tau_s$ or higher $D_{\rm max}$ parameter values.
%
We also propose theoretical distribution functions for both the lifespans ($\tau_s$) and the number of \cetis{} per unit time \citep{maccone_evolution_2014, Sotos_biotechnology_2019}.
%
The later is related to the time span between the appearance of consecutive \cetis{} (since when $\left<\tau_a\right>$ is shorter, it produces a greater density of \cetis{}).
%
The analytical expresiones for these distributions are set to a fixed law, as discussed in Sec.~\ref{SS_PDF_shape}.

@@

\begin{figure}[!t]
   \centering
   \includegraphics[width=0.5\textwidth]{F_scheme.pdf}
   \caption{
   %
Space--time diagrams showing a schematic representation of the
different stages in the development of causal contact node (\ceti{}).
%
Panel (a) represents the region $R$ in space and time which is causally
connected to the emitter (left vertex), following an "A" type event
("Awakening") in which the node acquires the communication capability.
%
The sphere of first contact (SFC) is a sphere centered on the emitter
that grows until its radius reaches the D$_{\rm max}$ distance, in which
the power of a signal would equal the detectability threshold.
%
In the Figure this sphere is represented by the left triangle of the
region $R$.
%
The surface of last contact (SLC) is another sphere that grows from a
"D" event ("Doomsday").
%
The region which is causally connected to the emitter is then limited
by these two spheres, and has the shape of a sphere or of a spherical
shell, depending on the time.
%
The temporal intervals for
the communication between two \cetis{} are represented in he
panel (b).
%
The receiver \ceti{} E$_1$ can listen signals from the emiter \ceti{} E$_2$,
from the "Contact" event (t=C$_{12}$) up to the "Blackout" event (t=B$_{12}$).
%
Similarly, 
the receiver \ceti{} E$_2$ can listen signals from the emiter \ceti{} E$_1$,
from t=C$_{21}$ up to t=B$_{21}$. 
%
   } \label{F_scheme}
\end{figure}

@@

\ffn{1} Space—time diagrams, where time and space are represented on
the horizontal and vertical axes, respectively, are suitable tools to
describe the causal connections among different nodes.
%
We illustrate in the Fig.~\ref{F_scheme} the schematic representation
of the region causally connected to a given \ceti{}.
%
For the sake of simplicity, we show in the plot the light travel
distance, i.e., the distance traversed by any signal spreading from
the emitter at light speed (for example, any electromagnetic signal).
%
In this scheme, a light pulse would follow a trajectory represented by
a line at 45 degrees from the axes, given the units of the time and
space axes are years and light years, respectively.
%
Panel (a) of the Figure represents the region $R$ (shaded polygon)
which is causally connected to the emitter (left vertex).
%
This region develops after an event (hereafter dubbed A-type event) in
which the node acquires the communication capability, or becomes
''active''.
%
The sphere of first contact (SFC) is centred on the emitter,
represented in the Figure by the left angle of the region $R$.
%
This sphere grows until its radius reaches the D$_{\rm max}$ distance, in
which it would be no longer detectable due to the decrease in the
energy per unit area, which falls under the assumed detectability
threshold.
%
Similarly, the surface of last contact (SLC) is another sphere that
grows from an event in which the nodes ceases to possess the
communication capability (D type event), and carries the last signal
produced by the emitter.
%
The region which is causally connected to the emitter is then limited
by these two spheres and therefore has the shape of a filled sphere
(before the D event) or of a spherical shell (after the D event).
%
The temporal intervals for the communication between two \cetis{} are
represented in the panel (b) of the Figure.
%
In this scheme, the entire lifetime of a \ceti{} can be represented by
a polygon, as shown in the Figure.
%
The polygon is limited in the spatial axis by the maximum distance of
the signal $D_{\rm max}$ and in the time axis by the time span between the
A-type event (concave vertex) and the D-type event (convex vertex).
%
This representation allows to visualize the important events that
result from the presence and communications of two \cetis{}.
%
The points in time where the \cetis{} acquire their communicating
capacity are dubbed ‘’Awakening’’ events ($A_1$ and $A_2$ in the
Figure).
%
Similarly, the instants in time where the \cetis{} lose their
communicating capacity are dubbed ‘’Doomsday’’ events ($D_1$ and $D_2$
in the Figure).
%
The points in space--time where the first contact is produced for each
one of the \cetis{}, are defined as ''Contact'' events, shown as $C_1$
and $C_2$.
%
Finally, the two points where the contact is lost for each one the
\cetis{}, are denominated ''Blackout'' events, presented as $B_1$ and
$B_2$ in the Figure.
%
The receiver node E$_1$ can listen signals from the emitter node
E$_2$, from the first contact event (t=C$_{12}$) until the last
one (t=B$_{12}$).
%
Similarly, the receiver node E$_2$ can listen signals from the emitter
node E$_1$, from t=C$_{21}$ up to t=B$_{21}$.
%
The time intervals for the open communication channel are determined
by the ''C'' and ''B'' type events.
%
It is noteworthy the fact that there is a time delay between the
contact events of the two \cetis{} involved in this analysis, and also
a time delay between the two blackout events.
%
Therefore, the time range when a bidirectional contact is possible
occurs between the maximum time of the contact points and the minimum
time between the blackout points.

@@

The leading front, or surface of first contact (SFC) grows from the
central \ceti{} until it reaches the maximum distance $D_{\rm max}$.

@@

This region will grow if the surface of first contact has not yet
reached the maximum distance $D_{\rm max}$, and will shrink after a
D-type event until the surface of last contact reaches $D_{\rm max}$, producing as a result the loss of all signal from the central \ceti{}.

@@

 \setlength{\tabcolsep}{10pt}
\begin{table*}
\centering
\begin{tabular}{cllll}
\hline
   \multicolumn{2}{l}{independent variables (hypothesis space)}
   &min. value&max. value&Nbins\\
\hline
   $\tau_{a}$ & Mean temporal separation between consecutive
   awakenings & 10$^2$~yr & 10$^6$~yr & 50\\ 
   $\tau_{s}$ & Mean lifetime of a \ceti{}
   & 10$^2$yr & 10$^6$~yr& 50\\ 
   D$_{\rm max}$ & Maximum reach of a message  & 30~pc & 10~kpc & 50 \\
\hline
   \multicolumn{2}{l}{fixed variables} & \multicolumn{2}{c}{assumptions} &value \\
\hline
   & statistical properties of all \cetis{} &equally distributed&&\\
   & Point process for the distribution in time & homogeneous &&\\
   f$_s$ & The scan of the sky & fully efficient&&1\\
   f$_p$ & panspermia or colonization &absent&&0\\
   & shape of the Galactic Habitable Zone & 2--dimensional ring &&\\
   R$_{GHZ}^{\rm min}$   & Inner radius of the GHZ     & & & 7~kpc\\
   R$_{GHZ}^{\rm max}$   & Outer radius of the GHZ     &&& 9~kpc\\
   t$_{\rm max}$ & Time span of the simulation  &&&1.e6 yr\\
    & \multicolumn{3}{l}{number of random realizations for each point in the hypothesis space} &50\\
\hline
   \multicolumn{2}{l}{discrete events} &
   \multicolumn{3}{l}{affected variables}\\
\hline
   A event & Awakening: a \ceti{} starts its communication
   capabilities &Number of active \cetis{}\\
   B event & Blackout: the end of the communication channel stops
   &Number causal channels\\
   C event & Contact: a new causal contact is produced &Number of causal channels\\
   D event & Doomsday: a \ceti{} ends its communication capabilities&Number of active \cetis{}\\
\hline

@@

The temporal structure of the process is defined by two distribution parameters. One of them represent the mean time
interval that a \ceti{} can emit and receive signals, and the other the mean time interval between the emergence of
consecutive \cetis{}. The spatial structure of the simulation is given by the size and shape of the Galactic Habitable
Zone and the maximum distance a signal can travel to be detected
(D$_{\rm max}$). The parameters for the temporal
distributions also determine the spatial properties, since the density of active \cetis{} in the galaxy depend on these
two parameters. For example, a small $\tau_a$ and a large $\tau_s$ will produce a galaxy densely populated with
\cetis{}. Also, some hypotheses regarding the shapes of the distributions of the temporal parameters must be made in
order to complete the simulation. \citet{forgan_spatiotemporal_2011} argues that the times at which different
civilisations become intelligent follow a Gaussian distribution, and then the distribution of inter-arrival times is an
inverse exponential. We assume that the distribution of the times of A events is a stationary Poisson process, and then
the distribution of the times between the appearance of new consecutive \cetis{} is exponential. Regarding the duration
of a \ceti{}, we propose that the distribution of the duration of \cetis{} is a stationary exponential distribution.
There are no clear arguments to conclude a statistical law for the later distribution, so that we propose it as a
working hipothesis.  A theoretical statistical distribution of the lifespan of civilisations would rely preferentially
on the basis of the underlying astrophysical and biological processes \citep{balbi_impact_2018}. For example,
\citet{maccone_lognormals_2014} argues that this distribution should be a log normal.

@@

\begin{figure} \centering
   \includegraphics[width=0.5\textwidth]{F_number_of_contacts.pdf}
   \caption{
   %
Empirical cumulative distributions of the number contacts
for \cetis in four different samples (M5, M6, M7, M8) with two
values for the range of the signal reach: D$_{\rm max}$=40000 ly (blue),  and 
D$_{\rm max}$=80000 ly (orange).
%
Except for the models corresponding to a densely populated galaxy with
long--living \cetis, the \cetis which receive more than 10 contacts
are very unlikely.
%
See Table \ref{T_selected_models} for model descriptions.
%
   } \label{F_number_of_contacts}
\end{figure}

@@

In this Section we discuss the degree of complexity in the model, considering a compromise between the accuracy of the
model and the number of parameters that are free or with a high uncertainty.
%
Firstly, we emphasise that the odds of a causal contact between two \cetis{} should not be considered as the odds of a
contact between two intelligent civilisations, and in fact the latter could be much lesser than the former.
%
Indeed, in order to establish a contact between any two entities, a minimum degree of compatibility must be accomplished
without any previous agreement, making the possibility of a contact with a message that could be deciphered highly rare
\citep[see e.g. ][]{forgan_collimated_2014}.
%
Besides the trade--off between the simplicity and the complexity of the experiments, further analysis could be performed
following this framework in order to explore possible implications of the results for more detailed configurations.
%
For example, the communication method (isotropic, collimated, serendipitous) can affect the observables, making it
necessary to implement a correction factor.
%
Taking this into account, our results regarding the probability of causal contact should be considered as upper limits
for effective contacts, since they depend on the efficiency of both the emitter and the receiver to broadcast and scan
the sky for intelligent signals, respectively \citep{grimaldi_signal_2017}.
%
A correction by coverage ratio in the detection and by a targeting ratio in the emission could be easily implemented in
the simulation, although the effect of reducing the probability of contact is basically the product of the efficiency
ratios and thus such implementation is not necessary.
%
Therefore, the values of the probabilities could be modified by a constant correction factor equal to the combined
emission/reception efficiency \citep{smith_broadcasting_2009, anchordoqui_upper_2019, forgan_collimated_2014} or for
beam-like transmissions \citep{grimaldi_signal_2017}. 
%
Other considerations include the effects of alignments, the use of stars as sources or amplifiers \citep{Edmondson2003,
borra_searching_2012}, the nature of the message carrier.
%
Another improvement could be the use of a spatial distribution that resembles the spiral shape and the width of the disc
of the Galaxy.
%
Regarding the distributions of the parameters, different distribution functions for the mean lifetime of \cetis{} can be
implemented as alternatives.
%
Finally, regarding the degree of idealisation, the model admits different efficiencies for the sphere of the causal
region of each \ceti{}, featuring different searching strategies \citep{hippke_interstellar_2017}. 
%
It is also possible to consider that $D_{\rm max}$ is different for different \cetis{}.
%
For example, a power law where powerful emissions are rare and low emissions are common could be an improvement to the
model.
%
In this work we chose not to implement this for the sake of simplicity.
%
Finally, the role of the message contents could influence on the lifespan of a \ceti{} that receives a message, although
the implementation of such behaviour would increase the number of free parameters and would be more speculative.

@@

A discrete event simulation is performed for a given model $M(\tau_a,
\tau_s, D_{\rm max})$ by keeping track of a set of variables that change each time an event happens. 
%
In addition, the model comprises elements that are fixed for all simulations, for example, the functional forms of the statistical distributions and the adopted values of particular variables (see Table XXX).
%
The main variables that follow the evolution of the simulation are: the positions of stars, which are sampled randomly within the GHZ; the time of the awakening of each \ceti{} (A event); and the time of the vanishing of each \ceti{} (D event). 
%
The variables that can be deduced from the previous ones include: the number of \cetis{} in casual contact with at least another \ceti{} at a given time; the number of \cetis{} as a function of time; the number of \cetis{} that receive a message at least one time; the number of \cetis{} that receive a message at least one time and successfully deliver an answer; and the number distribution of waiting time to receive a message.
%
All these quantities are updated each time one of the four events (A, B, C, D) occurs.

@@

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{F_never_contact.pdf}
   \caption{
   %
Rate of MPLs that never make contact (listening) as a
function of $\tau_a$ and $\tau_s$, for 
$D_{\rm max}$=10000 (left panel),
$D_{\rm max}$=40000 (middle panel), and
$D_{\rm max}$=80000 (right panel)
%
   }
   \label{F_never_contact}
\end{figure*}

\begin{figure*} % 2D color plot
   \centering
   \includegraphics[width=\textwidth]{F_C_at_A.pdf}
   \caption{
   %
Rate of MPLs that listen at the moment of the
awakening, as a funcion of $\tau_a$ and $\tau_s$, for
$D_{\rm max}$=10000 (left panel),
$D_{\rm max}$=40000 (middle panel), and
$D_{\rm max}$=80000 (right panel)
%
   }
   \label{F_C_at_A}
\end{figure*}

@@

% hearsay 
%
In order to make a reproducible project, we developed the tools
that allow to run and obtain the results shown in this work.
%
The simulations were implemented on a Python-3 code, dubbed \hs{},
which is publicly accessible through the GitHub
platform\footnote{https://github.com/mlares/hearsay} under the
MIT-license.
%
The project is in the process of registration with the ‘’Astrophysics
Source Code Library’’ \citep[ASCL, ][]{2015JORS....3E..15A,
2020ASPC..522..731A}
%
In order to achieve high performance we made use of vectorised methods
and array broadcasting properties provided by the NumPy package
\citep{oliphant_numpy_2006, vanderwalt_numpy_2011}
%
The combination of C-compiled Numpy routines for numerical
computations with a high-level language allows to obtain an easily
comprehensible and maintainable code.
%
From the viewpoint of a user, \hs{} is an object-oriented package that
exposes the main functionalities as classes and methods.
%
In the configuration step, the user prepares the set of simulation
parameters through an initialisation file.
%
Since configuration files and simulation results are persistent, it is
straightforward to keep track of different experiments and the
experiments can be revisited easily.
%
An in-depth description of the methods can be found in the
documentation, which is automatically generated from \hs{} docstrings
and made public in the read-the-docs service.

@@

\setlength{\tabcolsep}{10pt}
\begin{table*}
\centering
\begin{tabular}{ccccc}
\hline
   Model subset & D$_{\rm max}$ [lyr] & $\tau_a$ interval [yr] & $\tau_s$ interval 
   [yr]& description  \\
\hline
M1 & 40000 & [10000, 30000]   & [10000, 50000]   &dense, short lifetime\\
M2 & 40000 & [170000, 190000] & [10000, 50000]   &sparse, short lifetime\\
M3 & 40000 & [10000, 30000]   & [400000, 440000] &dense, long lifetime \\
M4 & 40000 & [170000, 190000] & [400000, 440000] &sparse, long lifetime\\
%
M5 & 80000 & [10000, 30000]   & [10000, 50000]   &dense, short lifetime\\
M6 & 80000 & [170000, 190000] & [10000, 50000]   &sparse, short lifetime\\
M7 & 80000 & [10000, 30000]   & [400000, 440000] &dense, long lifetime \\
M8 & 80000 & [170000, 190000] & [400000, 440000] &sparse, long lifetime\\
%
\hline
\end{tabular}
\caption{Selected models to analyze the behavior of simulation outputs
   and their dependence on simulation parameters.}
\label{T_selected_models}
\end{table*}

@@

We implemented the simulation of a regular grid of models varying over
the hypothesis space, which covers 15000 different models.
%
For each model, we simulated 50 realizations with different random
seeds, in order to perform a MonteCarlo estimation of the variance.
%
The parameters for the temporal aspects of the simulation (the mean
waiting time for the next awakening, $\left<\tau_a\right>$, and the
mean lifetime, $\left<\tau_s\right>$) cover the ranges 10$^2$-10$^6$
yr, with a linear partition of 50 values for each parameter.
%
For the D$_{\rm max}$ parameter, we take the range 30~pc to 10~kpc, on
50 linear bins.
%
This setup makes a total of 750000 simulations, each covering a time
range of ten million years.
%
\ttn{1}
%
In the Table \ref{T_simu_hypotheses} we show the three variable
parameters, the ranges of their values and the number of bins that
have been explored in the numerical experiments.
%
We also show the set of fixed parameters that take part in the simulation,
their values and the hypotheses that they represent.
